Left-linear rewriting in adhesive categories

### 2024-04-24 concur/fusione.tex

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

When dealing with left-linear rules, such a characterisation
fails. Leaving aside the formal details (the interested reader can
refer to \Cref{def:permcon}), the intuition is quite simple. In the
linear case one proves that two derivations using the same rules in
different orders are switch equivalent when the colimits of the two
derivations, seen as diagrams, are isomorphic and the isomorphism
properly commutes with the embeddings of the rules. When considering
left-linear derivations, this is no longer true. Intuitively this is
due to the fact that the application of a rule can merge parts of an
object thus making the colimit little informative.
%
This effect can be seen for derivations $\der{D}$ in
Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. They are not
abstraction equivalent but the colimit of both derivations is a single node with a self-loop, i.e.~the terminal
object in $\cat{Graph}$, in a way that  all commutativity requirements are
trivially satisfied.
% OLD VERSION
% This is readily seen by considering derivations $\der{D}$
% in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
% observed that $\der{D}$ and $\der{D}'$ are not switch equivalent.
% % 
% Still, it is easy to see that they are related by a consistent\todo{le definiamo solo nell'appendice}
% permutation, which is the identity on productions. The colimit of
% both derivations is a single node with a self-loop, i.e.~the
% terminal object in $\cat{Graph}$, hence all commutativity requirements
% are trivially satisfied.

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two
	non-empty decorated\todo{never introduced} derivations $(\der{D}, \alpha, \omega)$ and
	$(\der{D}', \alpha', \omega')$ with the same length and with
	isomorphic sources and targets.

	If $\der{D}=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ with associated sequence of rules
	$r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$, a \emph{consistent permutation}
	between $(\der{D}, \alpha, \omega)$ and
	$(\der{D}', \alpha', \omega')$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are,
	respectively, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$

	\[
	\xymatrix@C=30pt{\pi(G_0)\ar[r]^{\alpha} \ar[d]_{\alpha'} & G_0
		\ar[r]^{\iota_{G_0}} &\tpro{D} \ar[d]^{\xi_\sigma}& \pi(G_{n+1})
		\ar[r]^{\omega} \ar[d]_{\omega'} & G_{n+1}
		\ar@{>->}[r]^{\iota_{G_{n+1}}} &\tpro{D} \ar[d]^{\xi_\sigma}\\
		G'_0 \ar[rr]_{\iota'_{G'_0}} & &\lpro \der{D}' \rpro& G'_{n+1}
		\ar@{>->}[rr]_{\iota'_{G'_{n+1}}} & &\lpro \der{D}' \rpro\\L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}
We can now turn to the proofs of the results of \Cref{subsec:canonical}.

>>>>

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and 10to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

When dealing with left-linear rules, such a characterisation
fails. Leaving aside the formal details (the interested reader can
refer to \Cref{def:permcon}), the intuition is quite simple. In the
linear case one proves that two derivations using the same rules in
different orders are switch equivalent when the colimits of the two
derivations, seen as diagrams, are isomorphic and the isomorphism
properly commutes with the embeddings of the rules. When considering
left-linear derivations, this is no longer true. Intuitively this is
due to the fact that a rewriting steps can merge parts of an
object thus making the colimit little informative.
%
This effect can be seen for derivations $\der{D}$ in
Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. They are not
abstraction equivalent but the colimit of both derivations is a single node with a self-loop, i.e.~the terminal
object in $\cat{Graph}$, in a way that  all commutativity requirements are
trivially satisfied.  
% OLD VERSION
% This is readily seen by considering derivations $\der{D}$
% in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
% observed that $\der{D}$ and $\der{D}'$ are not switch equivalent.
% % 
% Still, it is easy to see that they are related by a consistent\todo{le definiamo solo nell'appendice}
% permutation, which is the identity on productions. The colimit of
% both derivations is a single node with a self-loop, i.e.~the
% terminal object in $\cat{Graph}$, hence all commutativity requirements
% are trivially satisfied.

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two derivations $\der{D}:=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}':=\{\dder{D}'_i\}_{i=0}^n$ with the same length and with isomorphic sources and targets.
	Let also $r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$ be the  associated sequence of rules. 
	A \emph{consistent permutation}
	between $\der{D}$ and
	$\der{D}'$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$.
	\[
	\xymatrix@C=30pt{L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}
We can now turn to the proofs of the results of \Cref{subsec:canonical}.
